---
title: "Computing MIDI Pitch Bends from Note Numbers with Fractional Parts"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Computing MIDI Pitch Bends from Note Numbers with Fractional Parts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mydough)
```

## Motivation

As a microtonal algorithmic composer, I would like to be able to specify
the pitch of a note via MIDI note numbers with a fractional part. For
example, I can specify middle C with MIDI note number 60, and middle C
sharp with MIDI note number 61.

But suppose I want to compose in a quarter-tone scale - 24 notes to the
octave. A convenient way to do this would be using MIDI note numbers
with fractional parts.

```{r}
(quarter_tone_scale <- seq(60, 72, by = 1/2))

```

This is, of course, the easy part. But suppose I want to play the music
on a synthesizer that accepts MIDI control messages. That's the subject
of the next section.

## MIDI, pitch bend, channels, omni mode and MIDI Polyphonic Expression

Nearly all MIDI-controllable synthesizers will accept *pitch bend*
messages. And nearly all of those will allow you to set the lower and
upper limits of the pitch bend in semitones. This corresponds to the
lower and upper limits of the pitch bend wheel.

When the wheel is in the middle, the pitch bend is zero semitones. When
it's at the top, it's the upper limit and when it's at the bottom, it's
the lower limit. For our purposes, we want the range between the lower
and upper limits to be as small as possible as long as it's at least one
semitone. A range of -2 semitones to +2 semitones is usually available
and may even be the default.
