---
title: "mydough MIDI Score Tables"
output: rmarkdown::html_vignette
bibliography: "REFERENCES.bib"
vignette: >
  %\VignetteIndexEntry{mydough MIDI Score Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mydough)
install_mido()
library(reticulate)
use_condaenv("mydough")

```

## Why a score table?

The ultimate output document from a composer, whether a fully-automated
process or a human using automated processes, is a *score* of some kind.
If the composition is intended for human performers, the score would
usually be some mix of conventional music notation and specialized
composition-dependent notations. But for compositions intended to be
rendeded by a computer or a collection of electronic instruments, the
most "standard" way of defining a composition is a MIDI file. [@Back1999]

A composer usually thinks of events and time in musical terms - beats, bars,
repeated sequences / arpeggios / loops, chords and so on. A MIDI file has a
very precise way of dealing with time that is not particularly
musician-friendly. See [@midoMidoMIDI], sections 3.10.7 and 3.10.8 for the
details.

The `mydough` *score table* (symbolically, a `score_table`) provides an
intermediate format convenient for both humans and automated processes, which
`mydough` can then automatically compile into a MIDI file. A `score_table` is a
`data.table::data.table` [@Dowle2023]. `data.table`s provide the rapid in-place
data manipulation that the compilation process requires.

## Columns of a `score_table`

-   track_number: integer. A MIDI file can have a number of tracks.
-   start_time: numeric, unit is beats
-   duration: numeric, unit is beats. `mydough` will compute the end time.
-   midi_channel: integer. On the synthesizer end, this is the channel
    that the synth pays attention to. Channel numbers range from 1 to
    16.
-   pitch: numeric. This is the pitch of the note specified as a MIDI note
    number, an integer between 0 and 127, plus a fraction, which denotes a 
    pitch that fraction of a semitone higher than the pitch of the MIDI note
    number.\
    \
    To orient yourself, middle C is note number 60 and A 440 Hz is note number
    69. So middle C sharp would be `pitch` 61 and the quarter-tone between C
    and C sharp would be `pitch` 60.5. `mydough` translates a pitch to a
    `note_on` message and a `pitch_bend` message if one is needed.
-   velocity_on: integer. The speed at which the (virtual) key is pressed
(0 - 127).
-   velocity_off: integer. The speed at which the (virtual) key is released
(0 - 127).

## The compilation process

The compilation process is as follows:

1. The compiler sorts the `score_table`, first by `track_number` and then by
`start_time`.
2. The compiler determines the number of tracks in the _score_table_ and uses
that number, along with parameters supplied to the compiler function call, to
create an empty MIDI file object.
3. The compiler computes new columns with parameters that will be part of the
MIDI messages generated for the score.
4. The compiler then processes the `score_table` a row at a time, creating

## Synthesizers, channels and pitch bends
\
    \
    Some synthesiers can be set to "omni" mode, where they listen to all
    channels, or MIDI Polyphonic Expression (MPE) mode, where they use
    multiple channels internally to perform different notes. This is important
    in dealing with polyphonic music with non-default pitch classes.
    note should be bent from its standard value. Details:

    1.  This value is only used for note events.

    2.  For note events, speicfy a value of `NA` whenever there is no
        pitch bend.

    3.  Synthesizers often have a setting for the range of a pitch bend.
        `mydough` assumes it is between -200 and +200 cents.

    4.  Pitch bends apply to all the notes on the channel! So if you
        have more than one note sounding at a time, for example, in a
        chord, they either have to be sent to different synthesizers
        listening on different channels, or to one synthesizer listening
        in "omni" mode.

## References
