---
title: "`mydough` MIDI Score Tables"
output: rmarkdown::html_vignette
bibliography: "../../inst/REFERENCES.bib"
vignette: >
  %\VignetteIndexEntry{`mydough` MIDI Score Tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mydough)
library(reticulate)
use_condaenv("mydough")
py_discover_config()

```

## Why a score table?

The ultimate output document from a composer, whether a fully-automated
process or a human using automated processes, is a *score* of some kind.
If the composition is intended for human performers, the score would
usually be some mix of conventional music notation and specialized
composition-dependent notations. But for compositions intended to be
rendeded by a computer or a collection of electronic instruments, the
most "standard" way of defining a composition is a MIDI file.
[@Back1999]

The *score table* (symbolically, a `score-table`) provides a format
convenient for both humans and automated processes, which `mydough` can
then automatically compile into a MIDI file.

## Format of a `score_table`

First of all, a `score_table` is a `data.table::data.table`
[@Dowle2023]. `data.table`s provide rapid in-place data manipulation
that the compilation process requires.

Second, a composer usually thinks of events and time in musical terms -
beats, bars, repeated sequences / arpeggios / loops, chords and so on. A
MIDI file has a very precise way of dealing with time that is not
particularly musician-friendly. See [@midoMidoMIDI], sections 3.10.7 and
3.10.8 for the details.

### Columns of a `score_table`

-   track_number: integer. A MIDI file can have a number of tracks.

-   event_type: character. What kind of event is this? (It's usually a
    "note".)

-   event_start_time: real, unit is beats

-   event_duration: real, unit is beats. `mydough` will compute the end
    time

-   midi_channel: integer. On the synthesizer end, this is the channel
    that the synth pays attention to. Channel numbers range from 1 to
    16.\
    \
    Some synthesiers can be set to "omni" mode, where they listen to all
    channels, or MIDI Polyphonic Expression (MPE) mode, where they use
    multiple channels internally to perform different notes. This will
    be important when we come to pitch bends.

-   tempo: real, beats per minute. This can be changed on a "tempo"
    event only. The value is ignored on "note" events.

-   note_number: integer. For a note event, this is the MIDI note
    number, a value between 0 and 127. To orient yourself, middle C is
    note number 60 and A 440 Hz is note number 69.

-   pitch_bend: real, unit is cents. This is how much up or down the
    note should be bent from its standard value. Details:

    1.  This value is only used for note events.

    2.  For note events, speicfy a value of `NA` whenever there is no
        pitch bend.

    3.  Synthesizers often have a setting for the range of a pitch bend.
        `mydough` assumes it is between -200 and +200 cents.

    4.  Pitch bends apply to all the notes on the channel! So if you
        have more than one note sounding at a time, for example, in a
        chord, they either have to be sent to different synthesizers
        listening on different channels, or to one synthesizer listening
        in "omni" mode.

-   velocity: integer. How hard to play the note (0 - 127).

## References
